---
title: "Climate data analysis for climate risk reduction with R"
author: " By Michale Gebrekiors"
date: "May, 2024"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Custom Print Methods}
output:
  knitr:::html_vignette:
    toc: yes
---

```{r include=FALSE}
library(rmarkdown)
library(knitr)
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = " ")
```
<br>
<br>

# Data import
* Let us import the daily rainfall data using read.csv( ) function.


```{r echo=TRUE}
Mydata<-read.csv("D:/ClimateRiskReduction/Data/ClimateData/RainFallData/Daily_Rainfall_Data_for_32_Met_Stations.csv",header=TRUE) 

```

We can see the head and tail of our daily data (1992 to 2016) of 32 stations 

```{r echo=TRUE}
head(Mydata,1)

```


```{r echo=TRUE}
tail(Mydata,1)

```


## Make a date format

Dates are usaully entered into R as character strings and changed into date format using as.Date(x, "input_format") function where x is the character date and input_format gives the appropriate format for reading the date.The default format for inputting dates in R is yyyy-mm-dd.In our daily rainfall dataset, date is coded as a character variable in mm/dd/yy format. We can change the character in col 1 of our data frame to date format using the following codes


```{r}
Sta <- "1992-01-01" # start date
End <- "2016-12-31" # end date
date <- seq( from=as.Date(Sta), to =as.Date(End), by="days")#generate dates by day
Mydata$date<-date # change col one to date format

```


if we see head of our data, column  one is changed to date format
```{r}
head(Mydata,1)# col one and row one

```
<br>
<br>

## make a time series 

* zoo package can be installed using install.packages ("zoo")
* and load it

```{r}
library(zoo)## an R-package for time series
```


```{r}
MyDataTS <- zoo(Mydata[,-1],Mydata[,1])# chaning to zoo class
str(MyDataTS)# see it structure 
```
We can see head of our time series

```{r}
head(time(MyDataTS))

```

see start and end time of our time series

```{r}
start(MyDataTS)# start date  of data series
end(MyDataTS)# end date of data series

```
see the names of the stations
```{r echo=TRUE}
names(MyDataTS)

```

As example we can see time series plots of three stations 
```{r}
plot(MyDataTS[,c("Shire","Wukro","MekelleAP")],col=c("red","blue","black"),xlab="Year",lwd=2)

```

Let's us go further by a selecting a single station for Wukro city

```{r}
Mystation<-MyDataTS[,"Wukro"]# select a single station
```


```{r}
plot(Mystation,main="Wukro daily rainfall", col="blue",xlab="Year",ylab= " mm",lwd=2)
```

can we calulate the average mean long-term (1992 to 2015) daily rainfall? yes. we can use the function mean in R.

```{r}
mean(Mystation)
```
Why NA when we try to calculate mean? because of the availability of missing values in our data. missing in R is represented by NA. What about if we add one argument in the mean function in R 

```{r}
mean(Mystation,na.rm=TRUE)# the NA is removed here
```

<br>
<br>

# Data cleaning 
##Identify Missing Values

Typical data quality problems can be revealed by basic data summaries

```{r}
summary(Mystation)

```

The daily rainfall  is missing for 741 days it is about a 8% of 
the data. Why? Is it due to raingauge failure in theses days ? Did the meteorlogical agency failed in collecing rainfall for these days ?  The perecent of missing can be found as follows

```{r}
n <-length(Mystation)# number of data
n

```

we can use  functions like is.na() to identify missing. it gives us logical  values TRUE  and  FALSE  which are  equivalent  to  the  numeric  values 1 and 0 and therefore the sum() and other functions can be used to obtain useful information about missing data. see the following:


```{r}
nm<- sum(is.na(Mystation))# number of missing data
nm
```


```{r}
pm<- (nm/n)*100 # percent of missing 
pm
```

when were the missing values? at the begining,end or in the middle of the recorded periods? we can see them with the following code dipicted in red points.


```{r}

plot(Mystation,  main  =  "Wukro daily rainfall", col="blue",lwd=2, ylab  =  "mm",xlab="years")
abline(h  =  120,  col  =  "black")
points(xy.coords(x  =  time(Mystation),  y  =  120,  recycle  =  T),pch  =  ifelse(is.na(Mystation),  "l",  ""),  col  =  "red")
grid()

```

<br>
<br>

## Decision on missing data

Whatever the reason for missing daily rainfall data, we should decide on the most acceptable action in our model. We have three options

* Do we select only  periods with continous data? or
* Do we convert the missing values to mean or to another values?
* Do we drop all the rows where our rainfall data is missing?

There are a lot of methods for handiling missing data in climatology and hydrology in which we will discuss it in another topic . In this Wukro city, we might decide to drop the data rows with missing data about daily rainfall since there are less than 10% .


<br>
<br>


# Converting daily to lower temporal scale
## To monthly totals

*we have a program saved as DailyTomonthly.R in the RCode directory.

```{r}
source("D:/ClimateRiskReduction/RCode/DailyTomonthly.R")

```

```{r}
m<-DailyTomonthly(Mystation,sum,na.rm = TRUE)

```


```{r}
M <- matrix(m, ncol=12, byrow=TRUE)
colnames(M) <- month.abb
rownames(M) <- unique(format(time(m), "%Y"))
boxplot(M,main="Monthly rainfall distributions for Wukro city",xlab="Months",ylab="mm")

```
<br>
<br>


```{r}
## Daily to annual
source("D:/ClimateRiskReduction/RCode/DailyToAnnual.R")

```




##To annual totals

```{r}
A<-DailyToAnnual(Mystation,sum,na.rm = TRUE)
plot(A,main="Average annual rainfall total for Wukro city",type="o",lwd=2,col="blue",xlab="years",ylab="mm")

```




